\begin{figure}
\setlength{\tabcolsep}{3pt}
\begin{tabular}{rclcl}
$g$ & $\in$ & $\Global \subseteq \VarName$ \\
$\tl$ & $\in$ & $\ThreadLocal \subseteq \VarName$ \\
$l$ & $\in$ & $\Local \subseteq \VarName$ \\
$x,y$ & $\in$ & $\Var = \Global \cup \ThreadLocal \cup \Local$ \\
$v$ &  $\in$ & $\Value$ \\
$\sigma$ & $\in$ & $\Store = \Var \rightarrow \Value$ \\
$\varsG$ & $\in$ & $\StoreGlobal = \Global \rightarrow \Value$ \\
$\varsTL$ & $\in$ & $\StoreThreadLocal = \ThreadLocal  \rightarrow \Value$ \\
$\varsL$ & $\in$ & $\StoreLocal = \Local \rightarrow \Value$ \\
$e, \phi, \psi, \rho$ & $\in$ & $\StateExpr = 2^{\Store}$ \\
$\alpha, \beta$ & $\in$ & $\TransExpr = 2^{(\Store,\Store)}$ \\
$\locExpr$ & $\in$ & $\LocalStateExpr = 2^{\StoreLocal}$ \\
$P$ & $\in$ & $\ProcName$ \\
$A$ & $\in$ & $\ActionName$ \\
$m$ & $\in$ & $\Mover = \{B,R,L,N\}$\\
$\actions$ & $\in$ & $\ActionName \rightarrow (\StateExpr, \TransExpr, \Mover)$ \\
$\procs$ & $\in$ & $\ProcName \rightarrow (\StateExpr, 2^\ThreadLocal, \StateExpr, \Stmt)$ \\
$\lins$ & $\in$ & $\LinearVar = 2^{\Global \cup \ThreadLocal}$ \\
$\ProcLins$ & $\in$ & $(\ActionName \cup \ProcName) \rightarrow (\LinearVar, \LinearVar)$ \\
$\ABlockAny$ & $\in$ & $\mathit{InsideABlock} ::= \ABlockInside \mid \ABlockOutside$ \\
$\RefinementAny$ & $\in$ & $\mathit{InsideRefinement} ::= \RefinementInside \mid \RefinementOutside$ \\
$\Refines$ & $\in$ & $\ProcName \pf \ActionName$ \\
$\AvailableActions$ & $\in$ & $2^\ActionName$ 
\end{tabular}
\caption{Definitions}
\label{fig:definitions}
\end{figure}

\begin{figure}
\setlength{\tabcolsep}{3pt}
\begin{tabular}{rclcl}
$\stmt \in \Stmt$ &::= & $\skipstmt \mid \assert{\locExpr} \mid \yield{e,\lins} \mid$ \\
                  & & $\call{A} \mid \call{P} \mid \async{P} \mid $\\
                  & & $\ablock{e,\lins}{\stmt} \mid \stmt;\;\stmt \mid$\\
                  & & $\ite{\locExpr}{\stmt}{\stmt} \mid$ \\
                  & & $\while{e,\alpha}{\locExpr}{\stmt}$ \\ 
$\StmtStack \in \mathit{StmtStack}$ &::= & $\stmt \mid (\varsL,\StmtStack) \mid \StmtStack;\stmt$ \\
$T \in \mathit{Thread}$ &::= &$(\varsTL, (\varsL, \StmtStack))$ \\
$\Prog \in \Program$ &::= & $\ProgCtxt[\varsG][\varsTL][\varsL][\stmt]$ \\
~\\
$\StmtCtxt \in \mathit{StmtCtxt}$ &::= &$[]_{Stmt} \mid \StmtCtxt;\stmt$ \\
$\StmtStackCtxt \in \mathit{StmtStackCtxt}$ &::= & $([]_{\Local}, \StmtCtxt) \mid (L,\StmtStackCtxt) \mid \StmtStackCtxt;\stmt$ \\
$\ThreadCtxt \in \mathit{ThreadCtxt}$ &::= &$([]_{\ThreadLocal}, ([]_{\Local}, \StmtCtxt)) \mid$ \\
 & &$([]_{\ThreadLocal}, (\varsL, \StmtStackCtxt))$ \\
$\YieldingThread \in \mathit{YieldingThread}$ &::= &$\ThreadCtxt[\varsTL][\varsL][\yield{e,\lins}]$ \\
$\ProgCtxt \in \mathit{ProgCtxt}$ &::= & $(\procs, \actions, \ProcLins, []_{\Global}, \YieldingThreads \cdot \ThreadCtxt \cdot \YieldingThreads)$
\end{tabular}
\caption{Syntax}
\label{fig:syntax}
\end{figure}

In this paper, we will formalize our verification method on a core concurrent programming language \civl.
In this section, we present the syntax and operational semantics of our language.
Figure~\ref{fig:definitions} contains basic definitions.
Figure~\ref{fig:syntax} uses these definitions to define the syntax of a (dynamically executing) \civl program.

Informally, a \civl program is comprised of threads each of which has a stack, global variables shared across threads, 
thread-local variables that are shared across procedures, and local variables inside a procedure.
The sets $\Global$, $\ThreadLocal$, and $\Local$ are the names of global, thread-local, and procedure-local variables 
respectively.
These sets are mutually disjoint and $\Var$ denotes their union.
There is an uninterpreted set $\Value$ of values that may be stored in these variables.
$\Store$, $\StoreGlobal$, $\StoreThreadLocal$, and $\StoreLocal$ are the set of all maps into $\Value$
from $\Var$, $\Global$, $\ThreadLocal$, and $\Local$ respectively.
We denote elements of $\Store$, $\StoreGlobal$, $\StoreThreadLocal$, and $\StoreLocal$
by $\sigma$, $\varsG$, $\varsTL$, and $\varsL$ respectively.
Formally, a \civl program is a tuple $(\procs, \actions, \ProcLins, G, \TS)$ with the following components.

\noindent
{\bf Procedures.}
$\procs$ maps a procedure name $P \in \ProcName$ to a tuple $(\phi, \mods, \psi, \stmt)$, 
where $\phi$ is the precondition of $P$, $\mods$ is the set of thread-local variables potentially modified by $P$, 
$\psi$ is the postcondition of $P$, and $\stmt$ is the body of $P$.
The predicates $\phi$ and $\psi$ should not refer to procedure-local variables.
Procedures do not have parameters; however, parameters may be modeled using thread-local variables.

\noindent
{\bf Actions.}
$\actions$ maps an action name $A \in \ActionName$ to a tuple $(\rho,\alpha,m)$.
Actions are used inside procedure bodies to access variables.
The predicate $\rho$ is the precondition and the predicate $\alpha$ is the transition relation of $A$;
$\rho$ is a predicate over $\Store$ and $\alpha$ is a predicate over $\Store \times \Store$.
The final component $m$ is one of four values in $\{B,R,L,N\}$;
it denotes the commutativity type of the action, $B$ for both mover, $R$ for right mover, $L$ for left mover, 
and $N$ for non mover~\cite{FlanaganFLQ08}. 

\noindent
{\bf Linear interfaces.}
$\ProcLins$ maps each procdure and action name in $\ProcName \cup \ActionName$ to a linear interface 
$(\lins,\lins')$, where each of $\lins$ and $\lins'$ is a set comprising global and thread-local variables.
The procedure (or action) requires that the set $\lins$ be available at the beginning of its invocation 
and makes available the set $\lins'$ at the end of its invocation; 
we refer to $\lins$ and $\lins'$ as the input and output linear permissions respectively.

\noindent
{\bf Global store.}
$\varsG$ is the global store of the program.

\noindent
{\bf Threads.}
$\TS$ is a sequence of threads.
Each thread $T$ in $\TS$ is a pair $(\varsTL, (\varsL, \StmtStack))$, where $\varsTL$ is the thread-local 
store and $(\varsL, \StmtStack)$ the continuation of $T$.  
The continuation $(\varsL, \StmtStack)$ comprises a procedure-local store $\varsL$ and a statement stack $\StmtStack$.
Informally, $\varsL$ is the local store at the top of the pending stack of computation $\StmtStack$ of the thread.
A statement stack $\StmtStack$ is recursively defined to be either a statement $\stmt$ 
or a continuation $(\varsL,\StmtStack)$ or the sequential composition $\stmt;\StmtStack$ of a statement $\stmt$ with 
a statement stack $\StmtStack$.

We now describe the statements in \civl.
The statement $\skipstmt$ has no side effect and is used as a marker in the formal operational semantics.
The statement $\assert{\locExpr}$ fails if $\locExpr$ is false and otherwise has no effect.
The statement $\yield{e,\lins}$ yields control to the scheduler which may pick any existing thread to execute.
The predicate $e$, called a stable predicate, is expected to be established by the executing thread and preserved across
execution of other threads subsequent to the yield.
Similarly, the linear permissions $\lins$ are expected to be available as long as the control of the thread
is at the yield.
The statement $\call{A}$ invokes the action $A$;
if $\actions(A) = (\rho, \alpha, m)$, then the predicate $\rho$ is expected to hold at the invocation and 
the store is updated according to the transition relation $\alpha$.
The statement $\call{P}$ invokes the procedure $P$;
if $\procs(P) = (\phi, \mods, \psi, \stmt)$, then the precondition $\phi$ is expected to hold at the invocation
and the statement $\stmt$ begins execution.
The statement $\async{P}$ creates a new thread that begins by invoking the procedure $P$.
The statement $\ablock{e,\lins}{\stmt}$ is a block of statements that may not have a yield inside it and consequently executes atomically;
the predicate $e$ is expected to hold and the permissions $\lins$ expected to be available at the beginning.
The statement $\stmt_1;\;\stmt_2$ executes $\stmt_1$ followed by $\stmt_2$.
The statement $\ite{\locExpr}{\stmt_1}{\stmt_2}$ executes $\stmt_1$ if $\locExpr$ is true and otherwise executes $\stmt_2$.
The statement $\while{e,\alpha}{\locExpr}{\stmt}$ executes $\stmt$ repeatedly until $\locExpr$ becomes false.
The expression $e$ is a loop invariant and the expression $\alpha$ is a loop summary for yield-free executions.

The operational semantics of a \civl program is formalized as a relation $\trans$ between a pair of programs.
The rules for the relation (see supplemental material)
extend the syntax of statements with an embedded context~\cite{WrightF94}.
A statement context $\StmtCtxt$ is either a context $[]_{\Stmt}$ or the sequential composition of a 
statement context and a statement.
Thus, each statement context has a unique context $[]_{\Stmt}$ inside it;
this context encodes the current position of the program execution.
The result of substituting a statement $\stmt$ inside the statement context $\StmtCtxt$ is another statement and is denoted by $\StmtCtxt[\stmt]$.
Similarly, we define a statement stack context $\StmtStackCtxt$ with two contexts embedded inside it, $[]_{\Local}$ for substituting a procedure-local store
and $[]_{\Stmt}$ for substituting a statement.
The result of substituting a procedure-local store $\varsL$ and a statement $\stmt$ inside $\StmtStackCtxt$ is a statement stack and is denoted by $\StmtStackCtxt[\varsL][\stmt]$.
A thread context $\ThreadCtxt$ has three contexts embedded inside it, $[]_{\ThreadLocal}$ for substituting a thread-local store, 
$[]_{\Local}$ for substituting a procedure-local store, and $[]_{\Stmt}$ for substituting a statement.
The result of substituting a thread-local store $\varsTL$, a procedure-local store $\varsL$, and a statement $\stmt$ inside $\ThreadCtxt$ 
is a thread and is denoted by $\ThreadCtxt[\varsTL][\varsL][\stmt]$.

Finally, the result of substituting a global store $\varsG$, a thread-local store $\varsTL$, a procedure-local store $\varsL$, and a statement $\stmt$ inside $\ProgCtxt$ 
is a program and is denoted by $\ProgCtxt[\varsG][\varsTL][\varsL][\stmt]$.
It is important to note that all except at most one thread in a program are yielding threads blocked at a yield statement;
such a thread is captured by the symbol $\YieldingThread$ in Figure~\ref{fig:syntax}.
The one (possibly) non-yielding thread is the only one allowed to execute until it reaches a yield statement and becomes a yielding thread.
At that point, all threads are yielding and any one can be picked for execution.

$\Prog$ is {\em unsafe\/} if there exists $\Prog'$ such that $\Prog \trans^* \Prog'$ and $\Prog'$ fails an assertion.
$\Prog$ is {\em safe}, denoted $\Safe(\Prog)$, iff $\Prog$ is not unsafe.
Define the predicate $\Yielding$ as follows:
\[
\begin{array}{lcl}
\Yielding(\Prog) & = & \mathit{let}\ \Prog = (\procs, \actions, \ProcLins, G, \TS)\ \mathit{in}\\
                 &   & ~~~~~~\forall T \in \TS.\ T \in \mathit{YieldingThread}.
\end{array}
\]
$\Prog$ is {\em responsive\/}, denoted $\Responsive(\Prog)$, iff every infinite execution from $\Prog$ satisfies $\Yielding$ infinitely often.


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "paper"
%%% End: 


