\section{A concurrent programming language}
\label{sec:language}

\begin{figure}
\setlength{\tabcolsep}{3pt}
\scriptsize{
\begin{tabular}{rclcl}
$g$ & $\in$ & $\Global \subseteq \VarName$ \\
$\tl$ & $\in$ & $\ThreadLocal \subseteq \VarName$ \\
$l$ & $\in$ & $\Local \subseteq \VarName$ \\
$x,y$ & $\in$ & $\Var = \Global \cup \ThreadLocal \cup \Local$ \\
$v$ &  $\in$ & $\Value$ \\
$\sigma$ & $\in$ & $\Store = \Var \rightarrow \Value$ \\
$\varsG$ & $\in$ & $\StoreGlobal = \Global \rightarrow \Value$ \\
$\varsTL$ & $\in$ & $\StoreThreadLocal = \ThreadLocal  \rightarrow \Value$ \\
$\varsL$ & $\in$ & $\StoreLocal = \Local \rightarrow \Value$ \\
$e, \phi, \psi, \rho, \locExpr$ & $\in$ & $\StateExpr = 2^{\Store}$ \\
$\alpha, \beta$ & $\in$ & $\TransExpr = 2^{(\Store,\Store)}$ \\
$P$ & $\in$ & $\ProcName$ \\
$A$ & $\in$ & $\ActionName$ \\
$m$ & $\in$ & $\Mover = \{B,R,L,N\}$\\
$\actions$ & $\in$ & $\ActionName \rightarrow (\StateExpr, \TransExpr, \Mover)$ \\
$\procs$ & $\in$ & $\ProcName \rightarrow (\StateExpr, 2^{\ThreadLocal}, \StateExpr, \Stmt)$ \\
$\lins$ & $\in$ & $\LinearVar = 2^{\Global \cup \ThreadLocal}$ \\
$\ProcLins$ & $\in$ & $(\ActionName \cup \ProcName) \rightarrow (\LinearVar, \LinearVar)$ \\
$\ABlockAny$ & $\in$ & $\mathit{InsideABlock} ::= \ABlockInside \mid \ABlockOutside$ \\
$\Refines$ & $\in$ & $\ProcName \pf \ActionName$ \\
$\HiddenVars$ & $\in$ & $2^\Global$ \\
$\HiddenActions$ & $\in$ & $2^\ActionName$ \\
$\HiddenProcs$ & $\in$ & $2^\ProcName$ \\
$\Perm$ & $\in$ & $\Value \rightarrow 2^\Value$
\end{tabular}
}
\caption{Definitions}
\label{fig:definitions}
\end{figure}

\begin{figure}
\setlength{\tabcolsep}{3pt}
\scriptsize{
\begin{tabular}{rclcl}
$\stmt \in \Stmt$ &::= & $\skipstmt \mid \yield{e,\lins} \mid \call{A} \mid \call{P} \mid$ \\
                  & & $\async{P} \mid \ablock{e,\lins}{\stmt} \mid \stmt;\;\stmt \mid$\\
                  & & $\ite{\locExpr}{\stmt}{\stmt} \mid \while{e,\alpha}{\locExpr}{\stmt}$ \vspace{6pt} \\
\end{tabular}
\begin{tabular}{rclcl}
$\Frame \in \mathit{Frame}$ &::= & $(P, \varsL, \stmt)$ \\
$T \in \mathit{Thread}$ &::= &$(\varsTL, \Frames)$ \\
$\StmtCtxt \in \mathit{StmtCtxt}$ &::= &$[]_{\Stmt} \mid \StmtCtxt;\stmt$ \\
$\FrameCtxt \in \mathit{FrameCtxt}$ &::= & $(P, []_{\StoreLocal}, \StmtCtxt)$ \\
$\ThreadCtxt \in \mathit{ThreadCtxt}$ &::= &$([]_{\StoreThreadLocal}, \FrameCtxt\cdot\Frames)$ \\
$\ProgCtxt \in \mathit{ProgCtxt}$ &::= & $(\procs, \actions, \ProcLins, []_{\StoreGlobal}, \TS \cdot \ThreadCtxt \cdot \TS)$ \\
$\Prog \in \Program$ &::= & $\ProgCtxt[\varsG][\varsTL][\varsL][\stmt]$ \vspace{6pt} \\
$\YieldingThread \in \mathit{YieldingThread}$ &::= &$(\varsTL, (P,\varsL,\StmtCtxt[\yield{e,\lins}])\cdot\Frames) \mid$ \\
                                              &::= &$(\varsTL, (P,\varsL,\StmtCtxt[\call{P'}])\cdot\Frames) \mid$ \\
                                              &::= &$(\varsTL, (P,\varsL,\skipstmt)\cdot\Frames)$ \\
$\YProgram$ &::= & $(\procs, \actions, \ProcLins, \varsG, \YieldingThreads \cdot T \cdot \YieldingThreads)$ \\
$\CProgram$ &::= & $(\procs, \actions, \ProcLins, \varsG, \YieldingThreads)$
\end{tabular}
}
\caption{Syntax}
\label{fig:syntax}
\end{figure}

\begin{figure}
\scriptsize{
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\ProgCtxt[\varsG][\varsTL][\varsL][s] = (\_, \actions, \_, \_, \_) \\
\actions \vdash (\MakeStore{\varsG}{\varsTL}{\varsL}, s) \trans (\MakeStore{\varsG'}{\varsTL'}{\varsL'}, s')
}
{\ProgCtxt[\varsG][\varsTL][\varsL][s] \trans \ProgCtxt[\varsG'][\varsTL'][\varsL'][s']}
\;(\textsc{Step})
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\ProgCtxt[\varsG][\varsTL][\varsL][s] = (\_, \actions, \_, \_, \_) \\\\
\actions \vdash (\MakeStore{\varsG}{\varsTL}{\varsL}, s) \trans \error
}
{\ProgCtxt[\varsG][\varsTL][\varsL][s]) \trans \error}
\;(\textsc{Fail})
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\procs(P) = (\phi, \mods, \psi,\stmt) \\
\ProcLins(P) = (\lins, \lins') \\
T' = (\varsTL, (P, \varsL, \yield{\phi,\lins};\stmt)) \\
\TS = \Vector{T_1} \cdot \ThreadCtxt[\varsTL][\varsL][\async{P}] \cdot \Vector{T_2} \\
\TS' = \Vector{T_1} \cdot \ThreadCtxt[\varsTL][\varsL][\skipstmt] \cdot \Vector{T_2}
}
{
(\procs, \actions, \ProcLins, \varsG, \TS)
\trans
(\procs, \actions, \ProcLins, \varsG, \TS' \cdot T')
}
\;(\textsc{Async})
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
T = (\varsTL, \epsilon) 
}
{(\procs, \actions, \ProcLins, \varsG, \Vector{T} \cdot T \cdot \Vector{T'}) \trans (\procs, \actions, \ProcLins, \varsG, \Vector{T} \cdot \Vector{T'})}
\;(\textsc{End})
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\procs(P) = (\phi, \mods, \psi,\stmt) \\
\ProcLins(P) = (\lins, \lins') \\
T = (\varsTL, (P', \varsL, \StmtCtxt[\call{P}]) \cdot \Frames) \\
T' = (\varsTL, (P, \varsL, \stmt) \cdot (P', \varsL, \StmtCtxt[\skipstmt]) \cdot \Frames)
}
{(\procs, \actions, \ProcLins, \varsG, \Vector{T} \cdot T \cdot \Vector{T'}) \trans (\procs, \actions, \ProcLins, \varsG, \Vector{T} \cdot T' \cdot \Vector{T'})}
\;{(\textsc{Call})}
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
T = (\varsTL, (P,\varsL,\skipstmt) \cdot \Frames) \\ T' = (\varsTL, \Frames)
}
{(\procs, \actions, \ProcLins, \varsG, \Vector{T} \cdot T \cdot \Vector{T'}) \trans (\procs, \actions, \ProcLins, \varsG, \Vector{T} \cdot T' \cdot \Vector{T'})}
\;(\textsc{Return})
$
%%%%%%%%%%%%%%%%%%%%
}
\caption{Operational semantics for program}
\label{fig:operational-semantics1}
\end{figure}

\begin{figure}
\scriptsize{
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\actions(A) = (\rho, \alpha, m) \\
\vars \not \in \rho
}
{
\actions \vdash (\vars, \call{A}) \trans \error
}
\;{(\textsc{Atomic-False})}
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\actions(A) = (\rho, \alpha, m) \\
\vars \in \rho \\
(\vars, \vars') \in \alpha
}
{
\actions \vdash (\vars, \call{A}) \trans (\vars',\skipstmt)
}
\;{(\textsc{Atomic-True})}
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\\
}
{\actions \vdash (\vars, \yield{e,\lins}) \trans (\vars, \skipstmt)}
\;{(\textsc{Yield})}
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
}
{\actions \vdash (\vars, \ablock{e,\lins}{\stmt}) \trans (\vars, \stmt)}
\;{(\textsc{AtomicBlock})}
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\\
}
{\actions \vdash (\vars, \skipstmt;\stmt) \trans (\vars, \stmt)}
\;{\;\;\;\;\;\;\;\;\;\;\;\;(\textsc{Seq})}
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\vars \not\in \locExpr
}
{\actions \vdash (\vars, \ite{\locExpr}{s_1}{s_2}) \trans (\vars, s_2)}
\;{(\textsc{If-False})}
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\vars \in \locExpr
}
{\actions \vdash (\vars, \ite{\locExpr}{s_1}{s_2}) \trans (\vars, s_1)}
\;{(\textsc{If-True})}
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\vars \not\in \locExpr \\
}
{\actions \vdash (\vars, \while{e,\alpha}{\locExpr}{s}) \trans (\vars, \skipstmt)}
\;{(\textsc{While-False})}
$
\medskip
%%%%%%%%%%%%%%%%%%%%
$
\inferrule
{
\vars \in \locExpr \\
\stmt' = \while{e,\alpha}{\locExpr}{\stmt}
}
{\actions \vdash (\vars, \stmt') \trans (\vars, \stmt;\stmt')}
\;{(\textsc{While-True})}
$
%%%%%%%%%%%%%%%%%%%%
}
\caption{Operational semantics for statements}
\label{fig:operational-semantics2}
\end{figure}

In this paper, we will formalize our verification method on a core concurrent programming language, \civl.
In this section, we present the syntax and operational semantics of our language.
Figure~\ref{fig:definitions} contains basic definitions.
Figure~\ref{fig:syntax} uses these definitions to define the syntax of a \civl program.
Informally, a \civl program is comprised of threads each of which has a stack, global variables shared across threads, 
thread-local variables that are shared across procedures, and local variables inside a procedure.
Thus, a program contains all information to represent not only the static program 
written by the programmer but also the dynamic state of the program as it executes.
The sets $\Global$, $\ThreadLocal$, and $\Local$ are the names of global, thread-local, and procedure-local variables 
respectively.
These sets are mutually disjoint and $\Var$ denotes their union.
There is an uninterpreted set $\Value$ of values that may be stored in these variables.
$\Store$, $\StoreGlobal$, $\StoreThreadLocal$, and $\StoreLocal$ are all sets of maps into $\Value$
from $\Var$, $\Global$, $\ThreadLocal$, and $\Local$ respectively.
We denote elements of $\Store$, $\StoreGlobal$, $\StoreThreadLocal$, and $\StoreLocal$
by $\sigma$, $\varsG$, $\varsTL$, and $\varsL$ respectively.
Formally, a \civl program is a tuple $(\procs, \actions, \ProcLins, G, \TS)$ with the following components.

\noindent
{\bf Procedures.}
$\procs$ maps a procedure name $P \in \ProcName$ to a tuple $(\phi, \mods, \psi, \stmt)$, 
where $\phi$ is the precondition of $P$, $\mods$ is the set of thread-local variables potentially modified by $P$, 
$\psi$ is the postcondition of $P$, and $\stmt$ is the body of $P$.
The predicates $\phi$ and $\psi$ cannot refer to procedure-local variables.
Procedures do not have parameters, but parameters may be modeled using thread-local variables.

\noindent
{\bf Actions.}
$\actions$ maps an action name $A \in \ActionName$ to a tuple $(\rho,\alpha,m)$.
Actions are used inside procedure bodies to access global and thread-local variables.
The predicate $\rho$ over $\Store$ is the gate and the predicate $\alpha$ over $\Store \times \Store$ 
is the transition relation of $A$.
If the gate $\rho$ does not hold on the current state, the program fails;
otherwise, the program makes progress by executing the action $A$ based on the transition relation $\alpha$.
Finally, $m$ is one of four values in $\{B,R,L,N\}$;
it denotes the commutativity type of the action, $B$ for both mover, $R$ for right mover, $L$ for left mover, 
and $N$ for non mover~\cite{FlanaganFLQ08}. 

\noindent
{\bf Linear interfaces.}
$\ProcLins$ maps each procedure and action name in $\ProcName \cup \ActionName$ to a linear interface~\cite{Wadler90lineartypes}
$(\lins,\lins')$, where each of $\lins$ and $\lins'$ is a set comprising global and thread-local variables.
To exploit linearity during verification, the programmer provides a function $\Perm$ from $\Value$ to $2^\Value$.
The set $\Perm(v)$ is the set of {\em permissions\/} associated with a value $v \in \Value$.
The \civl type checker (Section~\ref{sec:linearity}) computes a set of {\em available\/} linear variables at each control location such that
the multiset union of the permissions associated with the available variables is guaranteed to be a set, i.e., 
permissions are never duplicated among linear variables.
The linear interface $(\lins, \lins')$ of a procedure (or action) indicates the available sets
at its beginning ($\lins$) and end ($\lins'$), respectively.
Linear interfaces allow \civl to reason about each procedure separately.

\noindent
{\bf Global store.}
$\varsG$ is the global store of the program.

\noindent
{\bf Threads.}
$\TS$ is a sequence of threads.
Each thread $T$ in $\TS$ is a pair $(\varsTL, \Frames)$ where
$\varsTL$ is the thread-local store and $\Frames$ is a stack of frames comprising the continuation of $T$.  
Each frame $(P, \varsL, \stmt)$ in $\Frames$ comprises a procedure $P$, its procedure-local store $\varsL$, 
and a statement $\stmt$.

\subsection{Operational semantics}
\label{sec:operational-semantics}

Figures~\ref{fig:operational-semantics1} and~\ref{fig:operational-semantics2} together present the operational semantics as a relation 
$\trans$ between a pair of programs.
The rules for the relation extend the syntax of statements with an embedded context~\cite{WrightF94}.
A statement context $\StmtCtxt$ is either a context $[]_{\Stmt}$ or the sequential composition of a 
statement context and a statement.
Thus, each statement context has a unique context $[]_{\Stmt}$ inside it;
this context encodes the current position of the execution of a thread.
The result of substituting a statement $\stmt$ inside the statement context $\StmtCtxt$ 
is another statement and is denoted by $\StmtCtxt[\stmt]$.
Similarly, we define a frame context $\FrameCtxt$ with two contexts embedded inside it, 
$[]_{\StoreLocal}$ for substituting a procedure-local store
and $[]_{\Stmt}$ for substituting a statement.
The result of substituting a procedure-local store $\varsL$ and a statement $\stmt$ inside 
$\FrameCtxt$ is a frame and is denoted by $\FrameCtxt[\varsL][\stmt]$;
here, $\varsL$ is the procedure-local store for the procedure containing the currently executed statement $\stmt$.
A thread context $\ThreadCtxt$ has three contexts embedded inside it, $[]_{\StoreThreadLocal}$ for substituting a thread-local store, 
$[]_{\StoreLocal}$ for substituting a procedure-local store, and $[]_{\Stmt}$ for substituting a statement.
The result of substituting a thread-local store $\varsTL$, a procedure-local store $\varsL$, and a statement $\stmt$ inside $\ThreadCtxt$ 
is a thread and is denoted by $\ThreadCtxt[\varsTL][\varsL][\stmt]$.
Finally, the result of substituting a global store $\varsG$, a thread-local store $\varsTL$, a procedure-local store $\varsL$, and a statement $\stmt$ inside $\ProgCtxt$ is a program and is denoted by $\ProgCtxt[\varsG][\varsTL][\varsL][\stmt]$.

The operational semantics use the notation $\MakeStore{\varsG}{\varsTL}{\varsL}$ to denote the concatenation of the 
stores $\varsG$, $\varsTL$, and $\varsL$.
Rule \textsc{Step} in Figure~\ref{fig:operational-semantics1} allows the program $\ProgCtxt[\varsG][\varsTL][\varsL][\stmt]$ to move 
to $\ProgCtxt[\varsG'][\varsTL'][\varsL'][\stmt']$
if $(\MakeStore{\varsG}{\varsTL}{\varsL}, \stmt)$ can move to $(\MakeStore{\varsG'}{\varsTL'}{\varsL'}, \stmt)$ 
according to any rule in Figure~\ref{fig:operational-semantics2} except for rule \textsc{Atomic-False}.
Similarly, rule~\textsc{Fail} allows $\ProgCtxt[\varsG][\varsTL][\varsL][\stmt]$ to fail if $(\MakeStore{\varsG}{\varsTL}{\varsL}, \stmt)$
fails according to rule \textsc{Atomic-False}. 
Rule \textsc{Async} describes the creation of a thread via $\async{P}$.  
The new thread is added to the sequence of threads in the program.
Rule \textsc{End} describes the termination of a thread; the terminating thread is removed
from the sequence of threads in the program.
Rules \textsc{Call} and \textsc{Return} describe the call and return of a procedure via $\call{P}$.

The rules in Figure~\ref{fig:operational-semantics2} provide the operational semantics for the statements in \civl.
The statement $\skipstmt$ has no side effect and is used as a marker in the formal operational semantics.
The statement $\yield{e,\lins}$ contains a predicate $e$ and the linear permissions $\lins$.
The yield predicate $e$ is expected to be established by the executing thread and preserved by other threads;
the linear permissions $\lins$ are expected to be available as long as the control of the thread
is at the yield statement.
The yield statement is the mechanism by which the programmer specifies the cooperative semantics over which 
the refinement check is performed.
The statement $\call{A}$ invokes the action $A$.
Suppose $\actions(A) = (\rho, \alpha, m)$ and $\ProcLins(A) = (\lins, \lins')$.
If $\rho$ does not hold, the statement $\call{A}$ fails, 
otherwise the store is updated according to the transition relation $\alpha$.
The permissions in $\lins$ are expected to be available at the beginning
and the permissions in $\lins'$ are guaranteed to be available at the end.
The statement $\call{P}$ invokes the procedure $P$.
Suppose $\procs(P) = (\phi, \mods, \psi, \stmt)$ and $\ProcLins(P) = (\lins, \lins')$.
It is expected that the precondition $\phi$ holds and the permissions $\lins$ are available at the beginning.
Then, the statement $\stmt$ executes.
When $\stmt$ terminates, the postcondition $\psi$ holds and the permissions $\lins'$ are available.
The predicates $\phi$ and $\psi$ are expected to be preserved by other threads in the environment. 
The statement $\async{P}$ creates a new thread that begins by invoking the procedure $P$.
The statement $\ablock{e,\lins}{\stmt}$ is useful only for compactly encoding non-interference checks (Section~\ref{sec:refinement});
the predicate $e$ is expected to hold and the permissions $\lins$ expected to be available at the beginning of $\stmt$.
In \civl, yield predicates, preconditions, and postconditions are the three sources of location invariants~\cite{OwickiG76}.

Finally, we have the standard primitives for sequencing, conditional execution, and looping.
The statement $\stmt_1;\;\stmt_2$ executes $\stmt_1$ followed by $\stmt_2$.
The statement $\ite{\locExpr}{\stmt_1}{\stmt_2}$ executes $\stmt_1$ if $\locExpr$ is true and otherwise executes $\stmt_2$.
The statement $\while{e,\alpha}{\locExpr}{\stmt}$ executes $\stmt$ repeatedly until $\locExpr$ becomes false.
The expression $e$ is a loop invariant and the expression $\alpha$ is a summary for failure-free executions of the loop that do not go 
through a yield statement (Section~\ref{sec:refinement}).
The conditional expression $\locExpr$ in $\mathit{if}$ and $\mathit{while}$ statements cannot refer to global variables.

We formalize the preemptive operational semantics of \civl as a relation $\trans$ with domain $\Program$ and 
codomain $\Program \cup \{\error\}$.
We use four contexts---$\StmtCtxt$, $\FrameCtxt$, $\ThreadCtxt$, and $\ProgCtxt$---to formalize $\trans$
as a structured operational semantics~\cite{WrightF94}.
In this semantics, a thread is chosen nondeterministically to execute a single statement.
The statement $\skipstmt$ has no side effect.
The statement $\yield{e,\lins}$ contains a predicate $e$ and a set $\lins$ comprising global and thread-local variables.
The yield predicate $e$ is expected to be established by the executing thread and preserved by other threads;
the variables $\lins$ are expected to be available at the yield statement.
The yield statement is the mechanism by which the programmer specifies the cooperative semantics (described later) over which 
the refinement check is performed.
In \civl, yield predicates, preconditions, and postconditions are the three sources of location invariants~\cite{OwickiG76}.
The statement $\call{A}$ invokes the action $A$.
Suppose $\actions(A) = (\rho, \alpha, m)$.
If $\rho$ does not hold, the statement $\call{A}$ fails and the program evolves to $\error$, 
otherwise the store is updated according to the transition relation $\alpha$.
The statement $\call{P}$ invokes the procedure $P$.
Suppose $\procs(P) = (\phi, \mods, \psi, \stmt)$.
It is expected that the precondition $\phi$ holds, then statement $\stmt$ executes, and finally the postcondition $\psi$ holds.
The predicates $\phi$ and $\psi$ are expected to be preserved by other threads in the environment. 
The statement $\async{P}$ creates a new thread that begins by invoking the procedure $P$.
The statement $\ablock{e,\lins}{\stmt}$ is useful only for compactly encoding non-interference checks;
the predicate $e$ is expected to hold and the set $\lins$ is expected to be available at the beginning of $\stmt$.
Finally, we have the standard primitives for sequencing, conditional execution, and looping.

We now define the cooperative operational semantics induced by the presence of yield statements in the program.
A {\em yielding thread\/} ($\mathit{YieldingThread}$ in Figure~\ref{fig:syntax}) is one that is waiting to execute at a yield statement,
a call, or a return from a call.
A program in $\YProgram$ is one in which all threads except at most one thread are yielding threads.
A program in $\CProgram$ is one in which all threads are yielding threads.
Thus, we have $\CProgram \subseteq \YProgram \subseteq \Program$.
Let $\ytrans$ be the relation obtained by restricting the domain and codomain of $\trans$ to $\YProgram$.
In a $\ytrans$-execution, the one (possibly) non-yielding thread is the only one allowed to execute until it reaches 
a yield statement and becomes a yielding thread; at that point, all threads are yielding and any one can be picked for execution.
We define the relation $\ctrans$ with domain $\CProgram$ and codomain $\CProgram \cup \{\error\}$ as follows:
\begin{enumerate}
\item 
$\Prog \ctrans \Prog'$ if there is a $\ytrans$-execution from $\Prog$ to $\Prog'$ such that every program on the execution 
except for $\Prog$ and $\Prog'$ is not in $\CProgram$.
\item
$\Prog \ctrans \error$ if there is a $\ytrans$-execution from $\Prog$ to $\error$ such that every program on the execution 
except for $\Prog$ is not in $\CProgram$.
\end{enumerate}

Consider a program $\Prog \in \CProgram$.
We write $\Safe(\Prog)$ if it is not the case that $\Prog \trans \error$.
We write $\Safe^*(\Prog)$ if $\Safe(\Prog')$ for all $\Prog'$ such that $\Prog \trans^* \Prog'$.
We write $\CSafe(\Prog)$ if it is not the case that $\Prog \ctrans \error$.
We write $\CSafe^*(\Prog)$ if $\CSafe(\Prog')$ for all $\Prog'$ such that $\Prog \ctrans^* \Prog'$.
We write $\Cooperative(\Prog)$ if for each infinite $\ytrans$-execution $\Prog \ytrans \Prog_0 \ytrans \Prog_1 \ytrans \cdots$ 
starting from $\Prog$, there is some $i \geq 0$ such that $\Prog_i \in \CProgram$.
We write $\Cooperative^*(\Prog)$ if $\Cooperative(\Prog')$ for all $\Prog' \in \CProgram$ such that $\Prog \ctrans^* \Prog'$.

\subsection{Definitions}
\label{sec:definitions}
We conclude this section by presenting a few definitions of concepts that are used in Section~\ref{sec:verification} to formalize
our verification method.
A predicate $\rho$ is {\em independent\/} of a variable $x \in \Var$ if for all $\sigma \in \Store$
and $v \in \Value$, if $\sigma \in \rho$ then $\sigma[x \rightarrow v] \in \rho$.
A transition relation $\alpha$ is {\em independent\/} of a variable $x \in \Var$ if for all $\sigma,\sigma' \in \Store$
and $v,v' \in \Value$, if $(\sigma,\sigma') \in \alpha$ then $(\sigma[x \rightarrow v], \sigma(x \rightarrow v')) \in \alpha$.
We define $\accessVars(\rho)$ to be the smallest set $X \subseteq \Var$ such that $\rho$
is independent of $x$ for all $x \in \Var \setminus X$.
We also define $\accessVars(\alpha)$ similarly.
For any transition relation $\alpha$, let $\elim{\Local}.\ \alpha$ be the transition relation
\[\{(\MakeStore{\varsG}{\varsTL}{\varsL}, \MakeStore{\varsG'}{\varsTL'}{\varsL'}) \mid \exists \varsL_1,\varsL_2.\ (\MakeStore{\varsG}{\varsTL}{\varsL_1}, \MakeStore{\varsG'}{\varsTL'}{\varsL_2}) \in \alpha \}.\]
For any $X \subseteq \Var$, let $\Havoc(X)$, be the transition relation 
$\{(\sigma,\sigma') \mid \forall x \in \Var \setminus X.\ \sigma(x) = \sigma'(x)\}$.

Given $\rho_1, \rho_2 \in \StateExpr$, we write $\rho_1 \vee \rho_2$ for the union of $\rho_1$ and $\rho_2$, 
$\rho_1 \wedge \rho_2$ for the intersection of $\rho_1$ and $\rho_2$, and $\neg \rho_1$ for $\StateExpr \setminus \rho_1$.
Similarly, given $\alpha_1, \alpha_2 \in \TransExpr$, we write $\alpha_1 \vee \alpha_2$ for the union of $\alpha_1$ and $\alpha_2$, 
$\alpha_1 \wedge \alpha_2$ for the intersection of $\alpha_1$ and $\alpha_2$, and $\neg \alpha_1$ for $\TransExpr \setminus \alpha_1$.
We denote the relational composition of $\alpha_1$ and $\alpha_2$ by $\alpha_1 \circ \alpha_2$.
Given $\rho \in \StateExpr$ and $\alpha \in \TransExpr$, we denote by $\ga{\rho}{\alpha}$ the
set $\{(\sigma, \sigma') \mid \sigma \in \rho \Rightarrow (\sigma,\sigma') \in \alpha\}$.
We let $\rho \circ \alpha$ denote the relation obtained by restricting the domain of $\alpha$ to $\rho$.
Similarly, we let $\alpha \circ \rho$ denote the relation obtained by restricting the codomain of $\alpha$ to $\rho$.
We also sometimes use $\false$ to represent the empty set of states or the empty set of transitions.
Given a relation $R$ (over a pair of sets), we denote the domain of $R$ by $\dom(R)$ and the codomain of $R$ by $\cod(R)$.

It is often convenient to collect the permissions in a set of available variables.
To that end, we define the function $\Collect$.
Given a partial map $\pi$ from $\VarName$ to $\Value$ and a set $X \subseteq \VarName$,
let $\Collect(\pi,X)$ be the multiset $\bigcup \{\Perm(\pi(x)) \mid x \in X \wedge \dom(\pi)\}$.
For any multiset $\Lambda$ over $\Value$, we also define $\IsSet(\Lambda)$ be true iff $\Lambda$ is a set.
