\section{Overview}
\label{sec:overview}

\begin{figure}
\begin{verbatim}
var Color:[int]int; // WHITE=1, GRAY=2, BLACK=3
\end{verbatim}
\begin{verbatim}
procedure TopWB(linear tid:Tid, addr:int)
  atomic [if (Color[addr] == WHITE) { 
              Color[addr] := GRAY;  }];
{
  var cNoLock:int := GetColorNoLock();
  call YieldColorsGetDarker(); 
  if (cNoLock == WHITE) 
    call MidWB(tid, addr);
}
\end{verbatim}
\begin{verbatim}
procedure MidWB(linear tid:Tid, addr:int)
atomic [if (Color[addr] == WHITE) {
            Color[addr] := GRAY;  }];
{
  call AcquireLock(tid);
  var cLock:int := GetColorLocked(tid);
  if (cLock == WHITE) 
    call SetColorLocked(tid, GRAY);
  call ReleaseLock(tid);
}
\end{verbatim}
\begin{verbatim}
procedure YieldColorsGetDarker()
  ensures forall a:: Color[a] >= old(Color[a]);
{
  yield forall a:: Color[a] >= old(Color[a]);
}
\end{verbatim}
\begin{verbatim}
procedure AcquireLock() right [...];
procedure ReleaseLock() left [...];
procedure SetColorLocked() atomic [...];
procedure GetColorLocked() returns (cl:int)
  both [...];
\end{verbatim}
\caption{Program 5}
\label{fig:reft}
\end{figure}

{\bf Refinement.} 
\civl supports the verification of refinement of atomic action specifications for procedures. 
We illustrate this technique on Program 5 (Figure~\ref{fig:reft}),
a simplified version of a programming pattern in our concurrent garbage collector (GC).
Mutator threads entering a write barrier for an address \exC{addr} check
if the global variable \exC{Color[addr]} is \exC{WHITE}
and set it to \exC{GRAY}, indicating that the object at this address
and objects reachable from it should not be garbage collected. 

Procedure \exC{TopWB} implements the write barrier.
To avoid the cost of locking for each address encountered, \exC{TopWB} reads \exC{Color[addr]} without holding a lock.
If the color is {\tt WHITE}, it calls the more expensive procedure \exC{MidWB} 
that re-examines and possibly updates \exC{Color[addr]} while holding the lock.
\civl simplifies reasoning about \exC{TopWB} and \exC{MidWB} by allowing us to 
compactly express their specification as the following atomic action:
\begin{verbatim}
[if (Color[addr] == WHITE) Color[addr] := GRAY;]
\end{verbatim}
This specification indicates that regardless of the different implementations of 
\exC{TopWB} and \exC{MidWB} and regardless of how the environment interferes
with their execution, to their respective callers it appears as if they atomically execute the above code.

The correctness of \exC{topWB} is not obvious.
Consider the following potential scenario. 
\exC{TopWB}, not holding a lock, reads {\tt Color[addr]} and
sets \exC{cNoLock} to \exC{GRAY} and then yields inside the call to
\exC{YieldColorsGetDarker}. Another thread sets {\tt Color[addr]} to
{\tt WHITE}. \exC{TopWB} resumes, but does nothing and exits,
because the procedure-local variable \exC{cNoLock} is \exC{GRAY}. In this scenario, the atomic action
specification of \exC{TopWB} would not be satisfied. However, in the GC this
scenario is not possible. 
The yield predicate in \exC{YieldColorsGetDarker} expresses the fact that
other threads in the environment of the thread running \exC{TopWB} can
only modify {\tt Color[addr]} to a higher (darker) value. 
\civl then verifies that, for every control path through \exC{TopWB}
procedure, exactly one {\tt yield}-to-{\tt yield} execution
fragment implements the atomic action specification and that other fragments do not modify
global state. 
This proof requires both correct modeling of environment interference in \exC{YieldColorsGetDarker}
and the atomic action specification for the called procedure \exC{MidWB}.

There are two important benefits of atomic action specifications.
First, an atomic action is often the most compact and precise way to express the specification a procedure.
Second, an atomic action specification successfully hides from the caller of a procedure $P$
the potentially numerous yields inside the body of $P$.
This capability provides modularity akin to that provided by the frame rule for sequential programs.
Without this capability, any postcondition $\phi$ needed by the caller of $P$ must be made explicit in the precondition 
and yield predicates of $P$ and all procedures recursively called inside it.
With this capability, the caller achieves the same effect by writing $\yield{\phi}$ before and after the call, 
provided that $\phi$ is preserved by the atomic action specification of $P$.

{\bf Yield elimination.} 
We now turn our attention to the verification of \exC{MidWB} against
its atomic action specification.
Just as the verification of \exC{TopWB} builds on the specification of \exC{MidWB},
the verification of \exC{MidWB} builds on other refinement proofs (not shown) 
of the procedures called in \exC{MidWB};
these called procedures are shown at the bottom of the figure. 
Observe that in \exC{MidWB} there is no {\tt yield} between
statements. Since threads explicitly yield control, the
entire body of \exC{MidWB} is executed atomically. In
a real execution, control can switch between threads at any point in
the code. The absence of {\tt yield}s is justified by reasoning about
mover types and reduction. The procedures called in
\exC{MidWB} have the mover types claimed in their
declarations and verified by \civl. 
For example, the mover type of \exC{Acquirelock} is {\tt right} which indicates 
that it commutes later in time against concurrently executing environment actions.
\exC{ReleaseLock} has mover type {\tt left} and it commutes earlier in time;
\exC{SetColorLocked} has mover type {\tt atomic} and it does not commute;
\exC{GetColorLocked} has mover type {\tt both} and it commutes both earlier and later in time.
Given the mover types of 
atomic action specifications, the \civl verifier ensures
that the absence of yields is justified using reduction~\cite{Lipton75}.  
\civl then verifies that the body of \exC{MidWB}, without any yield statements, satisfies
its atomic specification. Yield elimination is useful both for the
user, since no yield predicates need to be provided, and for \civl,
as no predicates need to be checked for sequential correctness and
non-interference.

{\bf Variable hiding.} The atomic action specification of \exC{MidWB}
makes no reference to the lock variable, although its implementation
involves a lock. When verifying refinement for \exC{MidWB}, the lock
variable has been hidden. Hiding variables at refinement steps and
providing different {\tt yield} predicates at different phases of a
proof are important novel proof organization capabilities in \civl
that facilitate proofs spanning large abstraction gaps.

{\bf Putting it all together.} 
Refinement reasoning requires powerful and
flexible mechanisms for reasoning with invariants.  
As illustrated in Program~5, refinement verification for \exC{TopWB} required 
(1)~rely-guarantee reasoning using a yield predicate, and 
(2)~verification of atomicity for \exC{MidWB}, which, in turn, required reduction. 
The combination of features in \civl, such as assertion reasoning, linear variables, and yield elimination,
is designed to support the goal of organizing a
refinement proof in an efficient and natural way.
