\section{Modules}
\label{sec:modules}

\civl's features must scale up to larger programs in a modular way:
we should be able to check a large program by breaking it into smaller pieces and checking the pieces independently.
This section describes a simple module system built on \civl that allows separate verification of modules,
allowing programmers to make changes to the private implementation of one module
without disturbing the verification of other modules.

A key challenge for modular verification in \civl is the $\CommutativitySafe$ and $\InterferenceSafe$ judgments.
Naively, these are whole-program judgments,
quadratically checking all pairs of actions or all pairs of yields and atomic blocks from an entire program.
To achieve separate verification for \civl,
we must be able to check these judgments on a per-module basis.
To achieve this,
we observe that $\CommutativitySafe$ and $\InterferenceSafe$ are trivially true for operations that act on disjoint sets of global variables.
If an atomic block modifies only variables $g_1$ and $g_2$, it will not interfere with a yield that refers only to variables $g_3$ and $g_4$.
Suppose each module $M$ owns a set of global variables, such that each global variable is owned by exactly one module.
Then we can decree that only $M$'s procedures and actions can access $M$'s global variables.
Formally, define ownership for global variables, procedures and actions as:

\begin{tabular}{rclcl}
$M$ & $\in$ & $\Module$ \\
$\own_{\Global}$ & $\in$ & $\Global \rightarrow \Module$ \\
$\own_{\ProcName}$ & $\in$ & $\ProcName \rightarrow \Module$ \\
$\own_{\ActionName}$ & $\in$ & $\ActionName \rightarrow \Module$ \\
\end{tabular}

If $\own_{\ProcName}(P) = M$, then $P$'s preconditions, postconditions, and statements only refer to global $g$ where $\own_{\Global}(g) = M$.
Similarly, if $\own_{\ProcName}(A) = M$, then $A$'s gate and transition relation can depend only on $g$ where $\own_{\Global}(g) = M$.
With this restriction, $\CommutativitySafe$ and $\InterferenceSafe$ can be checked for each module $M$ in isolation;
no other module $M'$ could interfere with any of the global variables owned by $M$.

Nevertheless, a fixed ownership assignment is too inflexible to allow effective sharing of global state between modules.
Therefore, it's also important to notice that $\own_{\Global}$, $\own_{\ProcName}$, and $\own_{\ActionName}$ can change across a series of refinement layers.
This enables a style of modularity where a low-level module $M_1$ implements low-level actions on a set of global variables,
and then ownership passes from the low-level module $M_1$ to a higher-level module $M_2$ that builds on $M_1$.
This is particularly useful for libraries:
a library module implementing locks may define a variable to represent the abstract state of a lock;
after the lock module is verified at a low layer,
a client module can take ownership of the lock variable in a higher layer to aid verification of the client module
(see TODO TECH REPORT for a detailed example).
